Spitzer, Gaia and the Potential of the Milky Way

### 2013-07-24 text/paper1/paper.tex
change numbers to text, additions

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,;\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the computed enclosed mass
of the potential, $m_{enc}$, and the orbital radius of the satellite, $R$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=R\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,;\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}

Qualitatively, when the distance in this normalized 6D space,
$D_{{\rm ps},i}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
likely recaptured by the satellite, but this is not imposed as a
hard constraint. [JZSH ...] As a statistical measure of how many stars have recombined, we
construct the 6D distribution of the minimum phase-space vectors for
each star over each individual, integrated orbit. For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of this distribution of minimum distance vectors
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}(A_{ik})).
\end{equation}
[This objective function is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method..]

\subsection{Application to Simulated Data}
The LM10 simulation data (see Section \ref{sec:sgr}) is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{\url{www.astro.virginia.edu/~srm4n/Sgr/data.html}} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:} shows all 2D projections of the 6D minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential. Qualitatively, it is clear that the
variance of the distribution increases in the wrong potential. The objective function
(generalized variance) defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. Figure~\ref{fig:} shows [...]
curves produced by varying the true parameters of the potential by
10\% around the true value.

@@

>>>>

\subsection{The algorithm: Rewinder}
Quantifying this method requires a sample of stars with known full
space kinematics $(\bs{x}_{i}, \bs{v}_{i})|_{t=0}$ (e.g., measurements of all position and velocity components for these stars \emph{today} at $t=0$), the orbital parameters
for the progenitor system $(\bs{x}_p, \bs{v}_p)|_{t=0}$, and a
functional form for the potential, $\Phi({\boldsymbol\theta})$. For a
given set of potential parameters, $\boldsymbol\theta$, the orbits of the stars and
progenitor are integrated backwards for several Gigayears (determined from an
estimate of the interaction time of the system). The stars and
progenitor system are treated as test particles, so that the debris
stars do not become physically re-bound to the satellite. Instead, at each
timestep $t_j$, for each particle $i$, a set of normalized, relative
phase-space coordinates are
computed 
\begin{equation}
  \bs{q}_{i} = \frac{\bs{x}_{i} -
    \bs{x}_{p}}{R_{\rm tide}}\,\,\,\,;\,\,\,\,\bs{p}_{i} = \frac{\bs{v}_{i} -
    \bs{v}_{p}}{v_{\rm esc}}
\end{equation}
where $(\bs{x},\bs{v})_{i}$ and $(\bs{x},\bs{v})_{p}$ are the phase-space coordinates for the particles and progenitor, respectively. These definitions require an estimate of the mass of the
satellite, $m_{sat}$, which, combined with the orbital radius of the satellite, $R$, and the computed enclosed mass
of the potential within $R$, $m_{enc}$, sets the instantaneous tidal radius and
escape velocity,
\begin{equation}
  R_{\rm tide}=R\Big(\frac{m_{\rm sat}}{3m_{\rm enc}}\Big)^{1/3}\,\,\,\,;\,\,\,\,
  v_{\rm esc}=\sqrt{\frac{2Gm_{\rm sat}}{R_{\rm tide}}}.
\end{equation}
Qualitatively, when the distance in this normalized six-dimensional space,
$D_{{\rm ps},i}=\sqrt{|\bs{q}_{i}|^2+|\bs{p}_{i}|^2} < \sqrt{2}$, the star is
likely recaptured by the satellite. JZSH imposed a similar condition as a hard boundary and used the number of recaptured particles in a given backwards-integration as a cost function for optimization. What follows is a description of an updated procedure with a more natural choice for an objective function [TODO...].

As a statistical measure of how many stars have recombined, we
use the minimum phase-space vectors --- for
each star over the individual, integrated orbits --- as samples from an underlying six-dimensional distribution. For each star, $i$,
the phase-space distance, $D_{\rm ps}$, is computed at each timestep
$t_{j}$, and the vector with the minimum phase-space distance is stored
\begin{align}
  t^*_{i} &= \argmin_{t} D_{{\rm ps},i}\\
  \bs{A}_{i} &= (\bs{q}_{i}(t^*_{i}),\bs{p}_{i}(t^*_{i})).
\end{align}
Thus, the matrix $\bs{A}_{ik}$ contains these minimum phase-space distance vectors for each star,
where $k\in[1,6]$. Intuitively, the variance of this distribution of minimum distance vectors
will be larger for orbits integrated in an incorrect potential
relative to the distribution computed from the `true' orbital history
of the stars; in an incorrect potential, small differences in the
orbits of the stars relative to the orbit of the progenitor cause them
to spread out in this relative, normalized phase space. Thus, the
\emph{generalized variance} of the distribution, computed for a given
set of potential parameters, $\bs{\theta}$, is a natural choice for
the scalar objective function used in constraining the potential of
the Milky Way,
\begin{equation}
  f(\bs{\theta})=\det(\mathrm{Cov}( \bs{A}_{ik})).
\end{equation}

\subsection{Application to Simulated Data}
The LM10 simulation data (see Section \ref{sec:sgr}) is a perfect
test-bed for evaluating the effectiveness of this method. We start by
extracting both particle data and the satellite orbital parameters
from the present-day snapshot of the simulation
data.\footnote{\url{www.astro.virginia.edu/~srm4n/Sgr/data.html}} We then
`observe' a sample of 100 stars from the first leading and trailing
wraps of the stream: the radial velocity errors are drawn from a
Gaussian $\epsilon_{rv} \sim \mathcal{N}(\mu=0,\sigma=10~{\rm km/s})$,
the tangential velocity errors depend on the distance to the star and
are computed from the expected Gaia error curve (shown in
Figure~\ref{fig:gaia_errors}), and the distance errors on a star of distance
$D$ are $\epsilon_{dist} \sim \mathcal{N}(0,0.02\times
D)$. Figure~\ref{fig:variance6d} shows all two-dimensional projections of the six-dimensional minimum
phase-space distance distribution for the observed particles in the
correct and an incorrect potential: it is clear that the
variance of the distribution increases in the wrong potential. The 
generalized variance defines a convex hypersurface over which we
optimize four of the six logarithmic potential parameters:
$v_{circ}$, $\phi$, $q_1$, and $q_z$. We fix $q_2$ and $R_c$, which are degenerate with combinations of the other parameters. Figure~\ref{fig:} shows one-dimensional slices of the objective function
produced by varying each of the potential parameters by
$\pm20\%$ around the true values and holding all others fixed.

@@

% this needs to go somewhere down here...
[This objective function is an approximation to a more general likelihood function, described in detail in forthcoming work (Price-Whelan \& Johnston in prep.). The results shown in sections xx and yy are meant to illustrate the strength and concept of the method..]
